Equalizer Programming Guide

### 2008-05-05 Developer/ProgrammingGuide/paper.tex
Website: Add ohloh project metrics to download site
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.1 for Equalizer 0.5+svn\\[\medskipamount]
  \today
}

@@

Written by Stefan Eilemann.\\
Engineering contributions by Maxim Makhinya, Jonas B\"osch and Christian Marten.

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add statistics, improved images based on 1.2 print\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The \textsf{eqPly} example has a static distributed object to provide
initial data to all rendering nodes, as well as a versioned object to
provide frame-specific data such as the camera position to the
rendering methods.

\subsubsection{Object Usage for Rendering}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Distributed Data Handling}}
\end{wrapfigure}
Distributed objects are addressed using session-unique identifiers,
because pointers to other objects can not be distributed directly, since
they have no meaning on remote nodes. The session with which distributed
objects are registered is normally the \textsf{eq::Config}.

The shared data entry point for the render clients is the identifier
passed by the application to \textsf{Config::init}. This identifier
typically contains the identifier of a static distributed object, and is
passed by Equalizer to all \textsf{configInit} task methods. Normally
this initial object is mapped by the \textsf{eq::Node}, and contains
identifiers to other shared data objects.

@@

Applications distributing their scene graph, rather then relying on a
shared filesystem like \textsf{eqPly}, use the frame data also as an
entry point to their scene graph data structure. \fig{fObjects} shows
one possible implementation, where the identifier and version of the
scene graph root are transported using the frame data. The scene graph
root then serializes and de-serializes his immediate children by
transferring their identifier and current version.

The objects are still created by the application, and then registered or
mapped with the session in order to distribute them. When mapping
objects in a hierarchical data structure, their type often has to be
known in order to create them. Equalizer does not currently provide
object typing, this has to be done by the application, either implicitly
in the current implementation context, or by transferring a type
identifier. In \textsf{eqPly}, object typing is implicit since it is
well-defined which object is mapped in which task method.

@@

The way changes are to be handled is determined by calling
\textsf{Object::getChangeType} during the registration of the master
version of a distributed object . The change type determines the memory
footprint and the contract for calling the serialization methods. The
following change types are possible:

@@

{\footnotesize\begin{lstlisting}
void InitData::getInstanceData( eqNet::DataOStream& os )
{
    os << _frameDataID << _windowSystem << _useVBOs << _useGLSL << _filename;
}

void InitData::applyInstanceData( eqNet::DataIStream& is )
{
    is >> _frameDataID >> _windowSystem >> _useVBOs >> _useGLSL >> _filename;

@@

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd, VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

For each active render client, one \textsf{eq::Node} instance is
created on the appropriate machine. Nodes are only instantiated on their
render client processes, i.e., each process should only have one
instance of the \textsf{eq::Node} class. The application process might
also have a node class, which is handled in exactly the same way as the
render client nodes.

During node initialization the init data is mapped to a local instance
using the identifier passed from \textsf{Config::init}. The model is
loaded based on the filename in the initialization data. No pipe, window
or channel tasks methods are executed before \textsf{Node::configInit}
has returned.

{\footnotesize\begin{lstlisting}
bool Node::configInit( const uint32_t initID )
{
    eq::Config* config = getConfig();
    const bool  mapped = config->mapObject( &_initData, initID );
    EQASSERT( mapped );

    const string& filename = _initData.getFilename();
    EQINFO << "Loading model " << filename << endl;

    _model = new Model();
    if ( !_model->readFromFile( filename.c_str() ) )
    {
        EQWARN << "Can't load model: " << filename << endl;
        delete _model;
        _model = 0;
    }

    return eq::Node::configInit( initID );
}
\end{lstlisting}}%>>

The node config exit function deletes the loaded model and unmaps the
initialization data:

{\footnotesize\begin{lstlisting}
bool Node::configExit()
{
    delete _model;
    _model = NULL;

    eq::Config* config = getConfig();
    config->unmapObject( &_initData );

    return eq::Node::configExit();
}
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
bool Pipe::configInit( const uint32_t initID )
{
    const Node*     node        = static_cast<Node*>( getNode( ));
    const InitData& initData    = node->getInitData();
    const uint32_t  frameDataID = initData.getFrameDataID();
    eq::Config*     config      = getConfig();

@@

{\footnotesize\begin{lstlisting}
eq::WindowSystem Pipe::selectWindowSystem() const
{
    const Node*            node     = static_cast<Node*>( getNode( ));
    const InitData&        initData = node->getInitData();
    const eq::WindowSystem ws       = initData.getWindowSystem();

@@

\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes  (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

@@

        const Node*     node     = static_cast< const Node* >( getNode( ));
        const InitData& initData = node->getInitData();

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\small\label{fDB}Destination view of an DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

    Node*            node  = (Node*)getNode();
    const Model*     model = node->getModel();
    const eq::Range& range = getRange();

@@

In order to draw the model, a helper class for view frustum culling is
set up using the view frustum from Equalizer and the camera position
from the frame data. The frustum helper computes the six frustum planes
from the projection and modelView matrices. During rendering, the
bounding spheres of the model are tested against these planes to
determine the visibility with the frustum.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1.5em}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
  \vspace{-1.5em}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\small\label{fThreads}Threads within one node process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

The command thread processes Equalizer-related request from other nodes,
for example during \textsf{eqNet::Object} mapping. In some special cases
the command thread executes application code, for example when a remote
node maps a static or unbuffered object,
\textsf{Object::getInstanceData} is called from the command thread.

@@

Please note that the default thread synchronization synchronizes all
\textsf{Channel::frame\-Draw} operations on a \textbf{single} node with
the node's main thread. The per-node frame synchronization does not
break the asynchronous execution across nodes. Advanced applications can
remove the per-node frame synchronization.

@@

Each entity uses \textsf{waitFrameStarted} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::waitFrameStarted} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
At the end of the frame, two similar sets of synchronization methods are
used. The first set synchronizes the local execution, while the second
set synchronizes the global execution.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
texture-based volume rendering, where the 3D texture is intersected by
view-aligned slices. The slices are rendered back-to-front and blended
together to produce the final image, as shown in
\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
  Research Council (DFG)}.

@@

The full power of scalable volume rendering is however in DB (sort-last)
compounds, where the full volume is divided into separate bricks. Each
of the bricks is rendered like a separate volume. For recomposition, the
\textsf{RGBA} frame buffer data resulting from these render passes then
has to be assembled correctly.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

\begin{figure}[h!t]
  \subfigure[]{\label{fBlendOrtho}
    \includegraphics[width=.45\textwidth]{images/b2f_ortho.pdf}
  }\hfil
  \subfigure[]{\label{fBlendPerspective}
    \includegraphics[width=.55\textwidth]{images/b2f_perspective.pdf}
  }%
  {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
      Orthogonal and Perspective View Frustra}}
\end{figure}
For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

Finding the correct assembly order for perspective frustra is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frustra. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

@@

Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/statistics.png}
  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
The events are processed by the channel's and window's
\textsf{processEvent} method. The default implementation sends these
events to the config, using \textsf{Config::sendEvent}, as explained in
\sref{sEventHandling}. When the default implementation of
\textsf{Config::handleEvent} receives the statistics event, it sorts the
event per frame and per originator. When a frame has been finished, the
events are pushed to the local (app-)node for visualization.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.2 for Equalizer 0.5+svn\\[\medskipamount]
  \today
}

@@

Written by Stefan Eilemann.\\
Engineering contributions by Maxim Makhinya, Jonas B\"osch, Christian
Marten and Patrick Bouchaud.

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add statistics, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The \textsf{eqPly} example uses static distributed objects to provide
initial data and the model to all rendering nodes, as well as a
versioned object to provide frame-specific data such as the camera
position to the rendering methods.

\subsubsection{Common Usage}

Distributed objects are addressed using session-unique identifiers,
because pointers to other objects can not be distributed directly, since
they have no meaning on remote nodes. The session with which distributed
objects are registered is normally the \textsf{eq::Config}.

The shared data entry point for the render clients is the identifier
passed by the application to \textsf{Config::init}. This identifier
typically contains the identifier of a static distributed object, and is
passed by Equalizer to all \textsf{configInit} task methods. Normally
this initial object is mapped by the render clients, and it contains
identifiers to other shared data objects.

@@

@@

The way changes are to be handled is determined by calling
\textsf{Object::getChangeType} during the registration of the master
version of a distributed object. The change type determines the memory
footprint and the contract for calling the serialization methods. The
following change types are possible:

@@

{\footnotesize\begin{lstlisting}
void InitData::getInstanceData( eqNet::DataOStream& os )
{
    os << _frameDataID << _modelID << _windowSystem << _useVBOs << _useGLSL
       << _invFaces ;
}

void InitData::applyInstanceData( eqNet::DataIStream& is )
{
    is >> _frameDataID >> _modelID >> _windowSystem >> _useVBOs >> _useGLSL
       >> _invFaces;

@@

\subsubsection{\label{sSceneData}Scene Data}

Some applications might rely on a shared filesystem to access the data,
for example when out-of-core algorithms are used. Other applications
prefer to load the data only on the application process, and use
distributed objects to synchronize the scene data with the render
clients.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
\textsf{eqPly} uses static distributed objects to distribute the model
loaded by the application. This model can be easily extended to
versioned objects, in order to support dynamic data changes.

The kD-tree data structure and rendering code for the model is strongly
separated from Equalizer, and kept in the separate namespace
\textsf{mesh}. It could also be used in a GLUT application. To keep this
separation, an external 'mirror' hierarchy is constructed around the
tree. This hierarchy of \textsf{VertexBufferDist} nodes is responsible
for cloning the model data on the remote render clients.

The identifer of the root object of this distributed hierarchy is passed
as part of the \textsf{InitData}, and used during remote node
initialization to map the model. \fig{fModelDist} shows the UML
hierarchy of the model and distribution classes.

Each \textsf{VertexBufferDist} object has one node of the model's data
tree. It is serializing the data for this node. Furthermore, it mirrors
the kD-tree by having a \textsf{VertexBufferDist} child for each child
of its corresponding tree node. During serialization, the identifier of
these children is sent to the remote nodes, which reconstruct the mirror
distribution hierarchy and model data tree based on this data.

The serialization function \textsf{getInstanceData} sends all the data
needed to reconstruct the model tree: the object identifiers of its
children, vertex data for the tree root and vertex indices for the leaf
nodes, as well as the bounding sphere and database range of each
node. The deserialization function \textsf{applyInstanceData} retrieves
the data in multiple steps, and constructs the model tree on the fly
based on this information. It is omitted here for brevity:

{\footnotesize\begin{lstlisting}
void VertexBufferDist::getInstanceData( eqNet::DataOStream& os )
{
    EQASSERT( _node );
    os << _isRoot;

    if( _left && _right )
    {
        os << _left->getID() << _right->getID();

        if( _isRoot )
        {
            EQASSERT( _root );
            const mesh::VertexBufferData& data = _root->_data;

            os << data.vertices << data.colors << data.normals << data.indices;
        }
    }
    else
    {
        os << EQ_ID_INVALID << EQ_ID_INVALID;

        EQASSERT( dynamic_cast< const mesh::VertexBufferLeaf* >( _node ));
        const mesh::VertexBufferLeaf* leaf = 
            static_cast< const mesh::VertexBufferLeaf* >( _node );

        os << leaf->_vertexStart << leaf->_vertexLength << leaf->_indexStart
           << leaf->_indexLength;
    }

    os << _node->_boundingSphere << _node->_range;
}
\end{lstlisting}}%>>

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Distributed Data Handling}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

The objects are still created by the application, and then registered or
mapped with the session in order to distribute them. When mapping
objects in a hierarchical data structure, their type often has to be
known in order to create them. Equalizer does not currently provide
object typing, this has to be done by the application, either implicitly
in the current implementation context, or by transferring a type
identifier. In \textsf{eqPly}, object typing is implicit since it is
well-defined which object is mapped in which context.

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking devices, as this is an orthogonal
problem to parallel rendering. Tracking device support has already been
solved by a number of implementations\footnote{VRCO Trackd VRPN, etc.},
which can easily be integrated with Equalizer. The example code in
\textsf{eqPly} provides a reference implementation for the integration
of such a tracking library. \sref{sTracking} provides more background on
head tracking.

@@

For each active render client, one \textsf{eq::Node} instance is
created on the appropriate machine. Nodes are only instantiated on their
render client processes, i.e., each process should only have one
instance of the \textsf{eq::Node} class. The application process might
also have a node class, which is handled in exactly the same way as the
render client nodes. The application and render clients might use a
different node factory, instantiating a different type of \textsf{eq::Node}.

During node initialization the static, per-config data is mapped to a
local instance using the identifier passed from
\textsf{Config::init}. No pipe, window or channel tasks methods are
executed before \textsf{Node::configInit} has returned.

{\footnotesize\begin{lstlisting}
bool Node::configInit( const uint32_t initID )
{
    if( !eq::Node::configInit( initID ))
        return false;

    Config*    config = static_cast< Config* >( getConfig( ));
    return config->mapData( initID );
}
\end{lstlisting}}

The actual mapping of the static data is done by the config. The config
first retrieves the distributed \textsf{InitData}. The object is
directly unmapped since it is static, and the data has be retrieved.

The init data contains, among other things, the identifier of the
root of the model's kD-tree. Using this identifier, the model is mapped
to the node, as explained in \sref{sSceneData}:

{\footnotesize\begin{lstlisting}
bool Config::mapData( const uint32_t initDataID )
{
    if( _initData.getID() == EQ_ID_INVALID )
    {
        const bool mapped = mapObject( &_initData, initDataID );
        EQASSERT( mapped );
        unmapObject( &_initData ); // data was retrieved, unmap immediately
    }
    else  // appNode, _initData is registered already
        EQASSERT( _initData.getID() == initDataID );

    const uint32_t modelID = _initData.getModelID();
    if( modelID == EQ_ID_INVALID ) // no model loaded by application
        return true;

    if( !_model )
        _model = ModelDist::mapModel( this, modelID );

    return (_model != 0);
}
\end{lstlisting}}

Since the model data is static, the mirror tree for data distribution
is only created temporarily to retrieve the data. This is done in a
static function of the \textsf{VertexBufferDist} class. One instance of
the distribution class is mapped to the model's identifier. During
application of the instance data, it will map its children recursively:

{\footnotesize\begin{lstlisting}
mesh::VertexBufferRoot* VertexBufferDist::mapModel(eqNet::Session* session,
                                                   const uint32_t modelID )
{
    VertexBufferDist modelDist( 0, 0 );

    if( !session->mapObject( &modelDist, modelID ))
    {
        EQWARN << "Mapping of model failed" << endl;
        return 0;
    }

    modelDist._unmapTree(); // No longer needed

    return const_cast< mesh::VertexBufferRoot* >( modelDist._root );
}
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
bool Pipe::configInit( const uint32_t initID )
{
    const Config*     config    = static_cast<Config*>( getConfig( ));
    const InitData& initData    = config->getInitData();
    const uint32_t  frameDataID = initData.getFrameDataID();
    eq::Config*     config      = getConfig();

@@

{\footnotesize\begin{lstlisting}
eq::WindowSystem Pipe::selectWindowSystem() const
{
    const Config*            config = static_cast<Config*>( getConfig( ));
    const InitData&        initData = config->getInitData();
    const eq::WindowSystem ws       = initData.getWindowSystem();

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes  (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

@@

        const Config*   config   = static_cast< const Config* >( getConfig( ));
        const InitData& initData = config->getInitData();

@@

After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

    Config*          config = static_cast< Config* >( getConfig( ));
    const Model*     model  = config->getModel();
    const eq::Range& range  = getRange();

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\small\label{fDB}Destination view of a DB compound}}
\end{wrapfigure}
In order to draw the model, a helper class for view frustum culling is
set up using the view frustum from Equalizer and the camera position
from the frame data. The frustum helper computes the six frustum planes
from the projection and modelView matrices. During rendering, the
bounding spheres of the model are tested against these planes to
determine the visibility with the frustum.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\small\label{fThreads}Threads within one node process}}
\end{wrapfigure}
The command thread processes Equalizer-related request from other nodes,
for example during \textsf{eqNet::Object} mapping. In some special cases
the command thread executes application code, for example when a remote
node maps a static or unbuffered object,
\textsf{Object::getInstanceData} is called from the command thread.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
Please note that the default thread synchronization synchronizes all
\textsf{Channel::frame\-Draw} operations on a \textbf{single} node with
the node's main thread. The per-node frame synchronization does not
break the asynchronous execution across nodes. Advanced applications can
remove the per-node frame synchronization.

@@

Each entity uses \textsf{waitFrameStarted} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

At the end of the frame, two similar sets of synchronization methods are
used. The first set synchronizes the local execution, while the second
set synchronizes the global execution.

@@

The \textsf{eVolve} example is a scalable volume renderer. It uses 3D
texture-based volume rendering, where the 3D texture is intersected by
view-aligned slices. The slices are rendered back-to-front and blended
together to produce the final image, as shown in
\fig{fSlices}\footnote{Volume Data Set courtesy of: SFB-382 of the German
  Research Council (DFG)}.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
The full power of scalable volume rendering is however in DB (sort-last)
compounds, where the full volume is divided into separate bricks. Each
of the bricks is rendered like a separate volume. For recomposition, the
\textsf{RGBA} frame buffer data resulting from these render passes then
has to be assembled correctly.

@@

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

\begin{figure}[h!t]
  \subfigure[]{\label{fBlendOrtho}
    \includegraphics[width=.45\textwidth]{images/b2f_ortho.pdf}
  }\hfil
  \subfigure[]{\label{fBlendPerspective}
    \includegraphics[width=.55\textwidth]{images/b2f_perspective.pdf}
  }%
  {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
      Orthogonal and Perspective View Frustra}}
\end{figure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
Finding the correct assembly order for perspective frustra is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frustra. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/statistics.png}
  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

The events are processed by the channel's and window's
\textsf{processEvent} method. The default implementation sends these
events to the config, using \textsf{Config::sendEvent}, as explained in
\sref{sEventHandling}. When the default implementation of
\textsf{Config::handleEvent} receives the statistics event, it sorts the
event per frame and per originator. When a frame has been finished, the
events are pushed to the local (app-)node for visualization.
